---
title: "Assignment 5: Data Visualization"
author: "Daisy Chu"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
toc: true
---

## OVERVIEW

This assignment accompanies the Environmental Data Analytics module on
data visualization. You will create, format, combine, and interpret
plots using environmental datasets. Your goal is not only to produce
correct code, but also to make deliberate choices about themes, color,
accessibility, and the message communicated by each figure.

## Directions

1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing
    `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill
    each instruction.
4.  Be sure your code is tidy; use line breaks to ensure your code fits
    in the knitted output.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code
    into a single PDF file.
7.  All figures should include clear labels and units, readable text,
    and color choices appropriate for the type of variable displayed. Do
    not rely on color alone when another cue (such as shape, linetype,
    or facets) would make the figure more accessible.
8.  Complete your own code before using AI in the designated comparison
    activity in Question 7. Include the prompt and your evaluation of
    the AI response.

------------------------------------------------------------------------

## Set up your session

1.  Set up your session. Load the `tidyverse`, `here`, `cowplot`, and
    any color palette package(s) you choose to use. Verify your project
    directory. Read in the processed NTL-LTER nutrients and
    chemistry/physics file for Peter and Paul Lakes
    (`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`) and
    the processed Niwot Ridge litter file
    (`NEON_NIWO_Litter_mass_trap_Processed.csv`) from the
    `Processed_KEY` folder.

2.  Make sure R is reading dates as date format; if not change the
    format to date.

```{r initialize project, message=FALSE, warning=FALSE}
#1 
# load packages
library(tidyverse)
library(here)
library(cowplot)
library(RColorBrewer)
library(ggthemes)

# verify project directory
getwd()

# read in data sets
ntl_lter_peterpaul_processed <- read.csv(
  here("Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
)

neon_litter_processed <- read.csv(
  here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv")
)

#2 
# check date formats
class(ntl_lter_peterpaul_processed$sampledate)
class(neon_litter_processed$collectDate)

# change sampledate format to date
ntl_lter_peterpaul_processed <- ntl_lter_peterpaul_processed %>%
  mutate(sampledate = ymd(sampledate))

neon_litter_processed <- neon_litter_processed %>%
  mutate(collectDate = ymd(collectDate))

# check date formats again
class(ntl_lter_peterpaul_processed$sampledate)
class(neon_litter_processed$collectDate)

```

## Define your theme

3.  Build a theme and set it as your default theme. Start with a
    complete theme, such as `theme_minimal()` or `theme_classic()`, and
    then customize at least two of the following:

-   Plot background
-   Plot title
-   Axis labels
-   Axis ticks/gridlines
-   Legend

Choose a base font size that will remain readable when the plots are
knitted. Briefly identify the two or more elements you changed and
explain how the changes improve readability.

```{r build plot theme}
#3
# define a theme with bolded plot title and empty gridlines
my_theme <- theme_minimal(base_size = 11) + 
  theme(
    plot.title = element_text(face = "bold", size = 13),
    panel.grid.minor = element_blank()
  )

```

> Theme choices and rationale: I wanted to keep my plot theme simple, so I used
theme_minimal as my theme base, and added bold format for my main title. I also
wanted to remove the minor grid lines to minimize distraction in my plot.

## Create graphs

For Questions 4–7, create `ggplot2` graphs and adjust their aesthetics
using data-visualization best practices. Match the color palette to the
variable type, label variables and units clearly, and remove formatting
that does not support the comparison.

4.  [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`),
    using separate aesthetics for Peter and Paul lakes. Do not rely only
    on color: also use shape, linetype, facets, or another appropriate
    visual cue. Add a separate line of best fit for each lake using the
    `lm` method. Zoom in so that the central pattern is visible without
    deleting the extreme observations from the analysis (hint: use
    `coord_cartesian()` rather than `xlim()` or `ylim()`). Include a
    descriptive title, axis labels with units, and a meaningful legend
    title.

```{r plot total P vs PO4, message=FALSE, warning=FALSE}
#4 
# plot total phosphorous by phosphate for peter and paul lakes
tp_po4_peter_paul_plot <- ntl_lter_peterpaul_processed %>%
  ggplot(aes(x = po4, y = tp_ug, color = lakename)) +
  geom_point(size = 1, alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) + 
  facet_wrap(vars(lakename), nrow = 2) + 
  coord_cartesian(xlim = c(0, 43), ylim = c(0, 160)) +
  my_theme +
  labs(
    title = "Changes in Phosphorus Levels based on Phosphate Levels",
    subtitle = "Peter and Paul Lakes",
    x = "Phosphate levels (ug P/L)",
    y = "Total Phosphorus levels (ug P/L)",
    color = "Lake Name"
  )

tp_po4_peter_paul_plot

```

Question: What relationship do you observe between phosphate and total
phosphorus? Does the relationship appear to differ between lakes?
Identify any extreme observations and explain why zooming with
`coord_cartesian()` is preferable to removing them with scale limits.

> Answer: For Peter lake, as phosphate increases, total phosphorus
> increases. For Paul lake, as phosphate increases, total phosphorus
> increases very little and remains about the same. Paul lake had an
> extreme observation close to a phosphate level of 375. It could have
> been a data entry error, but because we are unsure, we don't want to
> discount it from our data set. By zooming in, we only visually remove
> the outlier out of frame to see the rest of the data better. However,
> the effects of the outlier still remain on our line of best fit, which
> we want rather than removing the outlier fully, using with scale
> limits, and eliminate how it impacts our data set.

5.  [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP,
    and (c) TN, with month on the x-axis and lake represented by a
    consistent aesthetic. Use a qualitative, color-vision-friendly
    palette. Then create a `cowplot` that combines the three graphs.
    Make sure that only one legend is present, the month order is
    correct, graph axes are aligned, and each y-axis states the variable
    and units.

Tips: \* Month should appear in calendar, not alphabetical, order. You
will need to convert to factor to order it. \* Setting an axis title in
your theme to `element_blank()` removes the axis title (useful when
multiple, aligned plots use the same axis values) \* Setting a legend's
position to "none" will remove the legend from a plot. \* Individual
plots can have different sizes when combined using `cowplot`.

```{r Create boxplots, fig.height=8, fig.width=8, warning=FALSE}
#5 
# make a boxplot of temperature
temp_boxplot <- ntl_lter_peterpaul_processed %>%
  mutate(month = sort(factor(format(sampledate, "%b"), levels = month.abb))) %>%
  ggplot(aes(x = month, y = temperature_C, color = lakename)) +
  geom_boxplot() +
  my_theme +
  theme(legend.position = "top") +
  labs(
    title = "Temperature, Total Phosphorus Levels, and Total Nitrogen Levels 
    Throughout the Months",
    subtitle = "Peter and Paul Lake",
    x = "Month",
    y = "Temperature (C)",
    color = "Lake Name"
  )

# make a boxplot of TP
tp_boxplot <- ntl_lter_peterpaul_processed %>%
  mutate(month = sort(factor(format(sampledate, "%b"), levels = month.abb))) %>%
  ggplot(aes(x = month, y = tp_ug, color = lakename)) +
  geom_boxplot() +
  my_theme +
  theme(legend.position = "none", element_blank()) +
  labs(
    x = "Month",
    y = "Total Phosphorus (ug P/L)"
  )

# make a boxplot of TN
tn_boxplot <- ntl_lter_peterpaul_processed %>%
  mutate(month = sort(factor(format(sampledate, "%b"), levels = month.abb))) %>%
  ggplot(aes(x = month, y = tn_ug, color = lakename)) +
  geom_boxplot() +
  my_theme +
  theme(legend.position = "none", element_blank()) +
  labs(
    x = "Month",
    y = "Total Nitrogen (mg N/L)"
  )

# create a cowplot to combine the three graphs
plot_grid(temp_boxplot, tp_boxplot, tn_boxplot, nrow = 3, align = 'v',
          rel_heights = c(1,0.6,0.6))
```

Question: What do you observe about the center, spread, and potential
outliers of each variable across seasons and between lakes? Support your
interpretation with specific features visible in the boxplots.

> Answer:For temperature, data is available only for Paul Lake in February and 
both lakes the months May through August. For total phosphorus and total 
nitrogen the data is present from June to November. For the center, which we
determine by looking at the center bar in each boxplot, we observe the highest 
centers for temperature of both lakes in May, before dropping down a little to 
around 10 degrees Celsuis during the summer months, with not much  difference 
in temperature between lakes. Peter lake seems to have a slightly larger spread 
of temperatures compared to Paul Lake. There don't seem to be any potential 
outliers for temperature, which we would have observed with circular points
outside of the ends of each boxplot.
The center for total phosphorus for Peter Lake is slightly higher than Paul Lake
across the summer and fall months. For the month of August, Paul Lake has a 
larger spread compared to Peter Lake. There seem to be outliers present for both
lakes in the months of June, August, September, October, and November.
The center for total nitrogen for Peter Lake in July is higher than the other
months. The spread of total nitrogen for Peter Lake is highest in July as well.
The spread of total nitrogen for Paul Lake is very minimal for the
mongths of Augst, September and November, but there are a lot of outliers 
observed for those months. We observe outliers for Peter Lake in all months
except for July.

6.  [Niwot Ridge] Subset the litter dataset to the `"Needles"`
    functional group. Plot dry mass of needle litter by date and
    distinguish NLCD classes using a qualitative color aesthetic. You do
    not need to rename the land-use classes. Format the date axis,
    labels, legend, point size, and transparency so that overlapping
    observations remain visible.

7.  [Niwot Ridge] First, independently create a second version of the
    plot with the three NLCD classes separated into facets rather than
    distinguished by color. Keep other formatting choices as consistent
    as possible so that you can compare the effect of faceting itself.

```{r Plot litter}
#6
# subset litter dataset to the Needles functional group
# and distinguish nlcd classes by color
needles_color <- neon_litter_processed %>%
  filter(functionalGroup == "Needles") %>%
  ggplot(aes(x = collectDate, y = dryMass, color = nlcdClass)) + 
  geom_point(size = 1.3, alpha = 0.5) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y",
               limits = as.Date(c("2016-01-01", "2019-12-31"))) +
  scale_color_discrete(labels = c("Evergreen Forest", "Grassland Herbaceous",
                                "Shrub Scrub")) +
  my_theme + 
  labs(
    title = "Changes in Dry Litter Mass of Needles Over Time",
    subtitle = "Classified by NLCD classes",
    x = "Date of Collection",
    y = "Dry Litter Mass (g)",
    color = "NLCD Class"
  )

needles_color

#7
# subset litter dataset to the Needles functional group
# and distinguish nlcd classes by facets
needles_facet <- neon_litter_processed %>%
  filter(functionalGroup == "Needles") %>%
  ggplot(aes(x = collectDate, y = dryMass, color = nlcdClass)) + 
  geom_point(size = 1.3, alpha = 0.5) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y",
               limits = as.Date(c("2016-01-01", "2019-12-31"))) +
  facet_wrap(vars(nlcdClass), nrow = 3, labeller = labeller(nlcdClass = c(
    "evergreenForest" = "Evergreen Forest",
    "grasslandHerbaceous" = "Grassland Herbaceous",
    "shrubScrub" = "Shrub Scrub"))) +
  scale_color_discrete(labels = c("Evergreen Forest", "Grassland Herbaceous",
                                "Shrub Scrub")) +
  my_theme +
  labs(
    title = "Changes in Dry Litter Mass of Needles Over Time",
    subtitle = "Classified by NLCD classes",
    x = "Date of Collection",
    y = "Dry Litter Mass (g)",
    color = "NLCD Class"
  )

needles_facet

```

## AI comparison

8.  After completing your own faceted plot, provide an AI tool with the
    Question 7 instructions and the relevant variable names. Ask it to
    propose `ggplot2` code for the same figure. Do **not** replace your
    original solution with the AI version.

Paste the exact prompt you used below:

> AI tool used: ChatGPT

> AI prompt: "The litter dataset has the name "neon_litter_processed". Subset 
 the litter dataset to the `"Needles"` functional group (variable name is 
 "functionalGroup"). Plot dry mass (variable name "dryMass") of needle litter by 
 date (variable name "collectDate") and plot with the three NLCD classes 
 (variable name "nlcdClass") separated into facets. You do not need to rename 
 the land-use classes. Format the date axis, labels, legend, point size, and 
 transparency so that overlapping observations remain visible. Include a title.
 Propose ggplot2 code for this visualization"

Run and inspect the AI-generated code, correcting only errors required
to make it run. If you make corrections, identify them. Then compare the
two solutions.

```{r AI solution}
# Paste the AI-generated code here. Clearly comment any corrections you make.
neon_litter_processed |>
  subset(functionalGroup == "Needles") |>
  ggplot(aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point(size = 2.5, alpha = 0.55,
             position = position_jitter(width = 2, height = 0)) +
  facet_wrap(~ nlcdClass) +
  labs(
    title = "Dry Mass of Needle Litter Over Time by NLCD Class",
    x = "Collection Date",
    y = "Dry Mass",
    color = "NLCD Class"
  ) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "2 months") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

9.  Which plot—your solution or the AI solution—more effectively
    communicates the data? Evaluate at least three specific choices,
    such as data filtering, facet structure, date-axis formatting,
    labels and units, theme, accessibility, or unnecessary code.
    Identify one useful idea from the AI solution and one choice you
    would reject or improve. If the solutions are very similar, explain
    why both approaches are appropriate.

> AI comparison: I believe my plot more effectively communicates the data. In
terms of facet structure, having each facet wrapped by row rather than
column made it easier to make comparisons of each NLCD class over time. Having
them side by side, like ChatGPT recommended, makes it more difficult to 
visually compare. The date-axis formatting by ChatGPT, in which date breaks
are by month, makes it impossible to see the month and year labels listed in
the x axis. Being able to experiment on my own and deciding to format date 
breaks by year and setting the date-axis bounds makes the plot simpler to read 
and understand. Lastly, there are too many grid lines that feel a bit 
unnecessary in ChatGPT's formatting. Manually implementing a theme that removes
minor grid lines helps create a cleaner aesthetic for my plot. I do like how
AI decided to format the legend to be at the bottom, so it doesn't take up
space from the plots themselves. The main choice I would change is the 
date-axis formatting AI decided to go with and opt for year date breaks instead
of month.

10. Question: Which encoding in Questions 6 and 7—color or facets—do you
    think is more effective for comparing NLCD classes, and why?
    Consider overlap, ease of comparison, accessibility, and the amount
    of space used by each figure.

> Answer: Classifying through facets is more effective for comparing NLCD
classes. Because there's no overlap of points in one plot, having three
separate plots makes it easier to disintguish patterns between NLCD classes.
Having three separate plots lined up on the same date-axis can be helpful
in visualizing comparisons. If someone is colorblind, it may be hard to 
distinguish between different NLCD classes just by color. Facets ensure that
each NLCD class can be easily distinguished without the use of color. Facet
wrapping does reduce the amount of space used by each figure, but because most
points are congregateded below 20g for dry litter mass, not a lot of vertical
space is needed for these plots.
