---
title: "Assignment 4: Data Wrangling"
author: "Daisy Chu"
output: pdf_document
geometry: margin=2.54cm
toc: true
---

## OVERVIEW

In this assignment, you will build a reproducible workflow that cleans four EPA
air-quality files, combines them, reshapes the result, and produces grouped
summaries. Date conversion is a brief review from Module 3; the emphasis is on
data wrangling.

## Directions

1. Rename this file `<FirstLast>_A04_DataWrangling.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.
6. Ensure that code in code chunks does not extend off the page in the PDF.

## Set up your session 

### Questions 1–2

1a. Load the `tidyverse` and `here` packages into your session.

1b. Check your working directory.

1c. Read all four raw files associated with the EPA Air dataset. See the README
file for information about the variables and measurements.

2. Add the appropriate code to reveal the dimensions of the four datasets. 

```{r setup questions 1-2, message = FALSE}
#1a load the tidyverse and here packages
library(tidyverse)
library(here)

#1b check working directory
getwd()

#1c read all four raw files associated with the EPA Air dataset
EPAair_O3_NC2024 <- read.csv(here("Data/Raw/EPAair_O3_NC2024_raw.csv"))
EPAair_O3_NC2025 <- read.csv(here("Data/Raw/EPAair_O3_NC2025_raw.csv"))
EPAair_PM25_NC2024 <- read.csv(here("Data/Raw/EPAair_PM25_NC2024_raw.csv"))
EPAair_PM25_NC2025 <- read.csv(here("Data/Raw/EPAair_PM25_NC2025_raw.csv"))

#2 see dimensions of the four data sets
dim(EPAair_O3_NC2024)
dim(EPAair_O3_NC2025)
dim(EPAair_PM25_NC2024)
dim(EPAair_PM25_NC2025)

```
The dimensions of each data set is listed below:
EPAair_O3_NC2024 has 9519 observations and 21 variables
EPAair_O3_NC2025 has 9466 observations and 21 variables
EPAair_PM25_NC2024 has 9752 observations and 22 variables
EPAair_PM25_NC2025 has 10184 observations and 22 variables


> All four datasets should have the same number of columns (PM2.5 has one extra column),
but unique record counts (rows). Do your datasets follow this pattern?
Yes, both of my ozone datasets have 21 columns and unique number of rows, while both
of my PM 2.5 data sets have 22 variables and a unique number of rows.

## Wrangle individual datasets to create processed files.

### Questions 3–5

3. Change the Date columns to be date objects.

4. Select the following columns: Date, Daily.AQI.Value, Local.Site.Name, 
AQS.Parameter.Description, County, Site.Latitude, Site.Longitude

5. For the PM2.5 datasets, fill all cells in AQS.Parameter.Description with "PM2.5" 
(all cells in this column should be identical).

```{r questions 3 thru 5}
#3 convert the Date columns of each data set to be date objects using mdy()
EPAair_O3_NC2024$Date <- mdy(EPAair_O3_NC2024$Date)
EPAair_O3_NC2025$Date <- mdy(EPAair_O3_NC2025$Date)
EPAair_PM25_NC2024$Date <- mdy(EPAair_PM25_NC2024$Date)
EPAair_PM25_NC2025$Date <- mdy(EPAair_PM25_NC2025$Date)

#4 select columns: Date, Daily.AQI.Value, Local.Site.Name, 
# AQS.Parameter.Description, County, Site.Latitude, Site.Longitude
# for each data set
EPAair_O3_NC2024_proc <- EPAair_O3_NC2024 %>%
  select(Date, Daily.AQI.Value, Local.Site.Name, AQS.Parameter.Description,
         County, Site.Latitude, Site.Longitude)

EPAair_O3_NC2025_proc <- EPAair_O3_NC2025 %>%
  select(Date, Daily.AQI.Value, Local.Site.Name, AQS.Parameter.Description,
         County, Site.Latitude, Site.Longitude)

EPAair_PM25_NC2024_proc <- EPAair_PM25_NC2024 %>%
  select(Date, Daily.AQI.Value, Local.Site.Name, AQS.Parameter.Description,
         County, Site.Latitude, Site.Longitude)

EPAair_PM25_NC2025_proc <- EPAair_PM25_NC2025 %>%
  select(Date, Daily.AQI.Value, Local.Site.Name, AQS.Parameter.Description,
         County, Site.Latitude, Site.Longitude)

#5 for PM2.5 data sets, fill all cells in the AQS.Parameter.Description
# column with "PM2.5"
EPAair_PM25_NC2024_proc$AQS.Parameter.Description <- "PM2.5"
EPAair_PM25_NC2025_proc$AQS.Parameter.Description <- "PM2.5"

```

### Question 6

6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r question 6}
#6 save all four processed data sets in the Processed folder
# rename files same as raw files but replace "raw" with "processed"
write.csv(EPAair_O3_NC2024_proc,
          row.names = FALSE,
          file = here("Data/Processed/EPAair_O3_NC2024_processed.csv"))

write.csv(EPAair_O3_NC2025_proc,
          row.names = FALSE,
          file = here("Data/Processed/EPAair_O3_NC2025_processed.csv"))

write.csv(EPAair_PM25_NC2024_proc,
          row.names = FALSE,
          file = here("Data/Processed/EPAair_PM25_NC2024_processed.csv"))

write.csv(EPAair_PM25_NC2025_proc,
          row.names = FALSE,
          file = here("Data/Processed/EPAair_PM25_NC2025_processed.csv"))

```


## Combine datasets

### Question 7

7. Stack the four datasets with `bind_rows()`. Check that their column names are
identical first.

```{r question 7}
#7 
# first check that the column names are identical
# grab column names from the EPAair_O3_NC2024_proc data set
colnames(EPAair_O3_NC2024_proc) == colnames(EPAair_O3_NC2025_proc) & 
  colnames(EPAair_O3_NC2024_proc) == colnames(EPAair_PM25_NC2024_proc) &
  colnames(EPAair_O3_NC2024_proc) == colnames(EPAair_PM25_NC2025_proc)

# columns have been confirmed to be identical
# stack the four data sets with bind_rows
EPAair_24_25 <- bind_rows(EPAair_O3_NC2024_proc, EPAair_O3_NC2025_proc,
                          EPAair_PM25_NC2024_proc, EPAair_PM25_NC2025_proc)

```

### Question 8

8. Use one pipeline (`%>%`) to create a daily site-level dataset that meets the
following conditions:

  * Include only sites that the four data frames have in common:  
    
  "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue",  
  "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", 
  "West Johnston Co.", "Garinger High School", "Castle Hayne", 
  "Pitt Agri. Center", "Bryson City", "Millbrook School" 
  
  (the function `intersect` can figure out common factor levels - but it will include sites with missing site information, which you don't want...) 

  * Some sites have multiple measurements per day. Group by date, site name,
  AQS parameter, and county, then calculate the mean AQI value, latitude, and
  longitude. Handle missing values explicitly in each mean.
  
  * Add numeric `Month` and `Year` columns from `Date`. Check if results are integers.
  
  * Hint: the dimensions of this dataset should be 15,012 × 9.

```{r question 8, message=FALSE, warning=FALSE}
#8 
# create a daily site-level data set
EPAair_daily <- EPAair_24_25 %>%
  # filter to the shared sites listed above
  filter(Local.Site.Name %in% c("Linville Falls", "Durham Armory", "Leggett", 
                                "Hattie Avenue", "Clemmons Middle", 
                                "Mendenhall School", "Frying Pan Mountain", 
                                "West Johnston Co.", "Garinger High School", 
                                "Castle Hayne", "Pitt Agri. Center", 
                                "Bryson City", "Millbrook School")) %>%
  # group by date, site name, AQS parameter, and county
  group_by(Date, Local.Site.Name, AQS.Parameter.Description, County) %>%
  # calc mean AQI value, latitude, and longitude
  summarise(mean_AQI = mean(Daily.AQI.Value),
            mean_latitude = mean(Site.Latitude),
            mean_longitude = mean(Site.Longitude)) %>%
  # add month and year columns from date
  mutate(Month = month(Date),
         Year = year(Date))

```

### Questions 9–10

9. Use `pivot_wider()` so that ozone and PM2.5 AQI values occupy separate
columns. Each location-date combination should occupy one row.

10. Call up the dimensions of your new tidy dataset.

```{r questions 9-10}
#9
# use pivot_wider() so that ozone and PM2.5 values occupy separate columns
EPAair_daily_wide <- EPAair_daily %>%
  pivot_wider(names_from = AQS.Parameter.Description,
              values_from = mean_AQI)

#10
# check the dimensions of the new data frame
dim(EPAair_daily_wide)

```
The dimension of the dataframe we pivoted wider is 9037 observations and
9 variables.

### Question 11

11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC2425_Processed.csv"

```{r question 11}
#11
# save processed data set to the filename: "EPAair_O3_PM25_NC2425_Processed.csv"
write.csv(EPAair_daily_wide,
          file = here("Data/Processed/EPAair_O3_PM25_NC2425_Processed.csv"))
```

## Generate summary tables

### Questions 12–13

12. Create a monthly summary data frame grouped by site, month, and year.
Calculate mean ozone and PM2.5 AQI for each group, handling missing values
explicitly. Then use `drop_na()` to remove rows only when mean **ozone** is
missing. Mean PM2.5 may remain missing.

13. Call up the dimensions of the summary dataset.

```{r questions 12 and 13}
#12
# create a monthly summary table with mean ozone and pm2.5 aqi values
EPAair_monthly_summary <- EPAair_daily_wide %>%
  group_by(Local.Site.Name, Month, Year) %>%
  summarize(mean_ozone = mean(Ozone),
            mean_pm2.5 = mean(PM2.5)) %>%
  drop_na(mean_ozone)

EPAair_monthly_summary_omit <- EPAair_daily_wide %>%
  group_by(Local.Site.Name, Month, Year) %>%
  summarize(mean_ozone = mean(Ozone),
            mean_pm2.5 = mean(PM2.5)) %>%
  na.omit()

#13
dim(EPAair_monthly_summary)
dim(EPAair_monthly_summary_omit)

```
The dimensions of the monthly summary data frame is 186 observations of 5 
variables.

The dimensions of the monthly summary data frame using `na.omit()` is 131
observations of 5 variables.

### Question 14

14. Why is `drop_na(mean_ozone)` more appropriate here than `na.omit()`? Test
both approaches and compare the dimensions of the resulting summary data frames.

> Answer: `drop_na(mean_ozone)` is more appropriate here than `na.omit()`
because the NAs for mean PM2.5 are clustered at certain sites such as 
Fyring Pan Mountain and Linville Falls. If we used na.omit and removed all NAs
instead of just mean ozone NAs, we would be omitting the majority of the data
rows for those sites. With just dropping the NAs for mean ozone, we can still
retain the information about the mean ozone values for those sites, rather than
losing all information altogether when dropping NAs for PM2.5 as well.
